Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada

### 2025-05-09 writing/manuscript.Rmd
shortened appendix; finished addressing comments

---
title: "Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'freshwater-biology.csl'
# csl: 'nature.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
    - \usepackage{booktabs} # to prevent tables from splitting
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

\noindent \textbf{References}: 73 (updated on 2025-01-16) <!-- counted by using nature csl -->

\noindent \textbf{Appendices}: 1

@@

* A review of climate change effects on marine mammals in United States waters: Past predictions, observed impacts, current research and conservation imperatives https://doi.org/10.1016/j.ecochg.2022.100054
* Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3
* boars' sleep quantity, efficiency and quality are reduced on warmer days (https://doi.org/10.1098/rspb.2023.2115)

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to maximize resource availability [e.g., food, water, shelter, see: @charnov_optimal_1976; @kacelnik_ideal_1992], optimize habitat use [@winter_forecasting_2024], and escape predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023]. Additionally, temperature also affects movement indirectly by through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010] and the value of resources [e.g., water, heat, shelter -- see: @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019]. Changes in temperature thus affects how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].

@@

\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)). ***MISSING INFO ON OTHER TELEMETRY DATA*** Clearly erroneous GPS locations were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behaviour. The script used to clean the data and all associated custom functions are available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).

@@

\noindent The following sections detail the statistical framework and modelling we used to estimate the effect on temperature on mammals' movement. To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the deviance explained, model predictions, and AIC values of the two sets of models.

keeping for thesis:

@@

\noindent In this study, we separate the effects of temperature on mammalian movement into indirect and direct effects. We call "indirect" all effects that can be intervened on by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022], and we refer to all other effects as "direct". For example, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can thus mitigate these indirect effects of temperature on movement behaviour by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability. In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behaviour (other than changing temperature directly). We provide the relevant Directed Acyclical Graphs in the Appendix (Figs. S1-S2).
-->

@@

\noindent We estimated the effects of temperature ($^\circ$C) on mammals' movement state (moving or not) and speed when moving using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019 and the code chunk below] with the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017]. The first HGAM estimated the probability that an animal was moving, $P(M)$, with a binomial family of distributions and logit link function. The response variable `moving` was coded as 1 if moving and 0 if not. The second HGAM estimated an animal's speed when moving (`speed`) with a gamma family of distributions and log link function. The HGAMs included random intercepts for each animal (`s(animal, bs = 're')`), fixed-effect intercepts for each species (`species`), and species-level `by` smooths (`s(..., by = species)`), which allowed each species' term to be estimated independent of other species [see model I in Figure 4 of @pedersen_hierarchical_2019]. The `by` smooths accounted for trends in time of day (in Pacific Daylight Time; `tod_pdt`), day of year (`doy`), and temperature (`temp_c`). Additionally, the models had three tensor product interaction terms (`ti()`) `by` each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year. These three terms accounted for: (1) seasonal changes in day length, (2) hourly changes in the response to temperature (e.g., changes in nocturnality), and (3) seasonal changes in the response to temperature (e.g., changes in coats and migration timing). Finally, smooth terms of log-transformed hours between GPS locations (`dt`) accounted for irregular sampling rates. A global term of `log(dt)` accounted for the overall effect of sampling interval, while a factor-smooth interaction term (`s(log(dt), species, bs = 'fs')`) allowed for species-level deviations from the global term while assuming a common smoothness parameter across species [see model GS in Figure 4 of @pedersen_hierarchical_2019]. The HGAMs accounted for the cyclicity of time of day and day of year using cyclic cubic splines [`bs = 'cc'`$;$ see p. 202 of @wood_generalized_2017]. Together, the binomial HGAM and the gamma HGAM inform us on an animal's long-term average speed, since it is the product of the probability of moving and its average speed when moving. We fit the models with fast REML (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optimize computational efficiency with no appreciable losses to model performance [@wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020]. Additional details are provided in Appendix A.

@@

intercepts <- tibble(Species = SPECIES,
                     'P(M)' = get_int(readRDS('models/binomial-gam-without-temperature.rds')),
                     'P(M|T)' = get_int(readRDS('models/binomial-gam.rds')),
                     'E(S)' = 'km/h' %#% get_int(readRDS('models/gamma-gam-without-temperature.rds')),
                     'E(S|T)' = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
                     'E(D)' = `P(M)` * `E(S)`,
                     'E(D|T)' = `P(M|T)` * `E(S|T)`) %>%
  mutate(Species = case_when(
    Species == 'Rangifer tarandus (s. mountain)' ~ 'R. tarandus (SM)',
    Species == 'Ursus arctos horribilis' ~ 'U. arctos horr.',
    Species == 'Puma concolor' ~ 'Puma concolor',
    Species == 'Cervus canadensis' ~ 'C. canadensis',
    Species == 'Rangifer tarandus (boreal)' ~ 'R. tarandus (boreal)',
    Species == 'Canis lupus' ~ 'Canis lupus',
    Species == 'Oreamnos americanus' ~ 'O. americanus')) %>%
  mutate(Species = paste0('\\emph{', Species, '}') %>%
           gsub(')}', ')', .) %>%
           gsub(' \\(', '} \\(', .))

@@

Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between +0.2% and +18.6%, with all HRSFs having an increase $\ge +4.6$%), and there was relatively strong agreement between models with and without temperature (Figs. S5 and S13), but including temperature resulted in substantially lower AIC values for all models, especially HRSFs (all $\Delta$AIC $\le$ -342; Table \@ref(tab:temp-comparison)). Although this may be interpreted as better fit, it is important to note that AIC generally decreases with more complex terms when sample sizes are large, even if the more complex model is not closer to the true model [see section 1.8.6, pages 52-54 of @wood_generalized_2017].

@@

temp_comp <- tibble(
  Model = c('$P(M)$', 'Speed', rep('HRSF', 7)),
  Species = c('All', 'All', sort(as.character(SPECIES))),
  DDE = 
    c(10.72 - 10.23, 79.26 - 79.11, 9.16, 4.62, 18.6, 8.0, 5.30, 12.0, 17.7) %>%
    round(1) %>%
    paste0('%'),
  DAIC = c(-5927, -1682, -6566, -4448, -813, -966, -5637, -342, -914) %>%
    round())

@@

knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of movement (P(moving)) and speed when moving (Speed) as well as in species-level Hierarchical Resource Selection Functions (HRSF).")
```

@@

```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the estimated effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{\\text{st}}$ (\\textbf{A}) and day of year at 12:00 PDT (\\textbf{B}). Predictions in the surface plots extend to 10% of the range away from each datum. The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to -75\\% and +300\\% (i.e., $2^{\\pm 2}$) for ease of readability."}
# not specifying dt, or doy for A and time for B because change is relative
knitr::include_graphics('../figures/distance.png')
```

@@

```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each location's average distance in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.8 and 1.25 ($\\approx 2^{\\pm 0.58}$; original data ranged 0.72 to 1.29). The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest cover."}
knitr::include_graphics('../figures/bc-displ-2100.png')
```

@@

```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in relative selection strength (RSS) for their current range, irrespective of RSS quantile or climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in RSS for the bottom 5\\%, median, and top 5\\% RSS values within the species' current extent. Changes are relative to the mean RSS for each quantile in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
knitr::include_graphics('../figures/rss-local-cc-predictions.png')
```

```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.6}$; original data ranged 0.26 to 117). The increase in boreal caribou's RSS for high elevations at warmer temperatures was highly uncertain and often not credible at the 95\\% level."}
knitr::include_graphics('../figures/bc-rss-2100.png')
```

@@

\noindent We have demonstrated that temperature is an important determinant of whether, how, and where mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources). Disentangling temperature from other drivers allowed us to predict how changes in climate will affect mammalian movement even if all other drivers remain approximately constant or adapt at negligible rates. However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are exposed to increasingly frequent conditions that are both extreme and novel. Predicting mammals' responses at the data-scarce fringes of the surface plots in Figs. \@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear. At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures. As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours, which will have the potential to substantially alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023]. For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [e.g., wild boars, *Sus scrofa*: @mortlock_sleep_2024] and likelihood to enter torpor [e.g., Australian eastern long-eared bat, *Nyctophilus bifax*: @fjelldal_nightly_2021]. Our results show that warmer temperatures cause grizzly bears to more further earlier in the year (Fig. \@ref(fig:dist)), which may alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024]. At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase their chances of finding food or being predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021] as well as the overall ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023]. These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011], including warmer temperatures and more frequent and intense extreme events [@yao_emergence_2022; @ipcc_climate_2023], such as heat waves, cold spells, floods and droughts, plant and animal die-offs, and sudden growth surges [sometimes referred to as "black swan" events -- see: @logares_black_2012; @anderson_black-swan_2017].

>>>>

---
title: "Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'movement-ecology.csl'
# csl: 'nature.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
    - \usepackage{booktabs} # to prevent tables from splitting
subparagraph: true # needed for \usepackage[small]{titlesec}
urlcolor: cyan
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

\noindent \textbf{References}: 140 (updated on 2025-05-09) <!-- counted by using movement ecology csl -->

\noindent \textbf{Appendices}: 2

@@

* Global vulnerability of marine mammals to global warming: https://doi.org/10.1038/s41598-019-57280-3

@@

\noindent For most animals, movement is a rapid and low-cost response to a variety of stimuli [@nathan_movement_2008]. Animals move to maximize resource availability [e.g., food, water, shelter, see: @charnov_optimal_1976; @kacelnik_ideal_1992], optimize habitat use [@winter_forecasting_2024], and escape predation [@peterson_pursuit_2021; @bartashevich_collective_2024; @tan_antipredator_2024]. However, ambient temperature affects mammals' movement rates by altering the energetic cost of movement [@mcnab_body_1970; @taylor_energetics_1982; @brown_toward_2004; @fuller_towards_2016; @jahn_variations_2022], the ease of movement [e.g., snow depth and density, see @montgomery_movement_2019; @leclerc_determinants_2021; @melin_effect_2023], and the risk of overheating [@hetem_responses_2014; @dyer_travel_2023]. Animals may search for a short-term buffer from extreme heat or cold via thermal refugia [@hannah_fine-grain_2014; @elmore_implications_2017; @attias_effects_2018; @arechavala-lopez_common_2019; @gulland_review_2022], which may even be preferred over forage abundance [e.g., @hall_microrefuges_2016]. Additionally, temperature also affects movement indirectly through other drivers of movement behaviour, such as trophic interactions [@hegel_interacting_2010]. Changes in temperature thus affects how animals move [e.g., movement frequency and speed -- their movement behaviour: @nathan_movement_2008; @dyer_travel_2023] and where they move to [i.e., their habitat selection: @merkle_large_2016; @alston_temperature_2020; @giroux_activity_2023].

@@

```{r, eval=FALSE}
tels <- readRDS('data/tracking-data/all-tracking-data-cleaned-2024-02-22-13-49.rds') %>%
  tidyr::unnest(tel)

# all outliers
sum(tels$outlier)
round(mean(tels$outlier) * 100, 2)

# new outliers
sum(tels$outlier & !tels$original_outliers)
round(mean(tels$outlier & !tels$original_outliers) * 100, 2)
```

\noindent Elk (*Cervus canadensis*) data from @ciuti_human_2012 were downloaded from Movebank [study name: Elk in southwestern Alberta, see @kays_movebank_2022], while boreal caribou (*Rangifer tarandus*) and wolf (*Canis lupus*) telemetries were acquired via a public BC Oil and Gas Research and Innovation Society repository ([https://www.bcogris.ca/projects/boreal-caribou-telemetry-data](https://www.bcogris.ca/projects/boreal-caribou-telemetry-data)). ***MISSING INFO ON OTHER TELEMETRY DATA.*** From the full set of telemetry data, a total of 2396 GPS locations (0.16%, including collar calibration data) were removed using diagnostic plots of (1) distance from the median location, (2) straight-line displacement between locations, (3) turning angle, and (4) time interval between consecutive points. Particular attention was paid to points with large turning angles ($\gtrapprox 170^\circ$) and excessively high straight-line displacement, especially if antecedent and subsequent points indicated stationary behaviour. See Appendix A for additional information.

@@

\noindent The following sections detail the statistical framework and modelling we used to estimate the effect on temperature on mammals' movement. To assess the importance of including temperature as an explicit covariate (rather than including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the deviance explained, model predictions, and AIC values of the two sets of models. We provide the relevant Directed Acyclical Graphs in the Appendix (Figs. S1-S2).

@@

\noindent In this study, we separate the effects of temperature on mammalian movement into indirect and direct effects. We call "indirect" all effects that can be intervened on by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022], and we refer to all other effects as "direct". For example, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can thus mitigate these indirect effects of temperature on movement behaviour by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability. In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behaviour (other than changing temperature directly). We provide the relevant Directed Acyclical Graphs in the Appendix (Figs. S1-S2).

@@

\noindent We estimated the effects of temperature ($^\circ$C) on mammals' movement state (moving or not) and speed when moving using two Hierarchical Generalized Additive Models [HGAMs, see @pedersen_hierarchical_2019 and the code chunk below] with the `mgcv` package for `R` [version 1.9-1, @wood_generalized_2017]. The first HGAM estimated the probability that an animal was moving, $P(M)$, with a binomial family of distributions and logit link function. The response variable `moving` was coded as 1 if moving and 0 if not. The second HGAM estimated an animal's speed when moving (`speed`) with a gamma family of distributions and log link function. The HGAMs included random intercepts for each animal (`s(animal, bs = 're')`), fixed-effect intercepts for each species (`species`), and species-level `by` smooths (`s(..., by = species)`), which allowed each species' term to be estimated independent of other species [see model I in Figure 4 of @pedersen_hierarchical_2019]. The `by` smooths accounted for trends in time of day (in Pacific Daylight Time; `tod_pdt`), day of year (`doy`), and temperature (`temp_c`). Additionally, the models had three tensor product interaction terms (`ti()`) `by` each species: (1) day of year and time of day, (2) temperature and time of day, and (3) temperature and day of year. These three terms accounted for: (1) seasonal changes in day length, (2) hourly changes in the response to temperature (e.g., changes in nocturnality), and (3) seasonal changes in the response to temperature (e.g., changes in coats and migration timing). Finally, smooth terms of log-transformed hours between GPS locations (`dt`) accounted for irregular sampling rates. A global term of `log(dt)` accounted for the overall effect of sampling interval, while a factor-smooth interaction term (`s(log(dt), species, bs = 'fs')`) allowed for species-level deviations from the global term while assuming a common smoothness parameter across species [see model GS in Figure 4 of @pedersen_hierarchical_2019]. The HGAMs accounted for the cyclicity of time of day and day of year using cyclic cubic splines [`bs = 'cc'`$;$ see p. 202 of @wood_generalized_2017]. Together, the binomial HGAM and the gamma HGAM inform us on an animal's long-term average speed, since it is the product of the probability of moving and its average speed when moving. We fit the models with fast Restricted Maximum Likelihood (`'fREML'`) and discretized covariates (`discrete = TRUE`) to optimize computational efficiency with no appreciable losses to model performance [@wood_generalized_2015; @wood_generalized_2017-1; @li_faster_2020]. Additional details are provided in Appendix B.

@@

intercepts <- tibble(Species = SPECIES,
                     'P(M)' = get_int(readRDS('models/binomial-gam-without-temperature.rds')),
                     'P(M|T)' = get_int(readRDS('models/binomial-gam.rds')),
                     'E(S)' = 'km/h' %#% get_int(readRDS('models/gamma-gam-without-temperature.rds')),
                     'E(S|T)' = 'km/h' %#% get_int(readRDS('models/gamma-gam.rds')),
                     'E(D)' = `P(M)` * `E(S)`,
                     'E(D|T)' = `P(M|T)` * `E(S|T)`) %>%
  mutate(Species = case_when(
    Species == 'Rangifer tarandus (s. mountain)' ~ 'R. tarandus (sm)',
    Species == 'Ursus arctos horribilis' ~ 'U. arctos horr.',
    Species == 'Puma concolor' ~ 'Puma concolor',
    Species == 'Cervus canadensis' ~ 'C. canadensis',
    Species == 'Rangifer tarandus (boreal)' ~ 'R. tarandus (b)',
    Species == 'Canis lupus' ~ 'Canis lupus',
    Species == 'Oreamnos americanus' ~ 'O. americanus')) %>%
  mutate(Species = paste0('\\emph{', Species, '}') %>%
           gsub(')}', ')', .) %>%
           gsub(' \\(', '} \\(', .))

@@

Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between +0.2% and +18.6%, with all HRSFs having an increase $\ge +4.6$%), and there was relatively strong agreement between models with and without temperature (Figs. S5 and S13), but including temperature resulted in substantially lower AIC values for all models, especially HRSFs (all $\Delta$AIC $\ge$ 342; Table \@ref(tab:temp-comparison)). Although this can be interpreted as evidence for a better fit, it is important to note that the addition of terms generally decreases AIC when sample sizes are large, even if the more complex model is not closer to the true model [see section 1.8.6, pages 52-54 of @wood_generalized_2017].

@@

temp_comp <- tibble(
  Model = c('$P(M)$', 'Speed', rep('HRSF', 7)),
  Species = c('All', 'All', sort(as.character(SPECIES))),
  DDE = 
    c(10.72 - 10.23, 79.26 - 79.11, 9.16, 4.62, 18.6, 8.0, 5.30, 12.0, 17.7) %>%
    round(1) %>%
    paste0('%'),
  DAIC = c(5927, 1682, 6566, 4448, 813, 966, 5637, 342, 914) %>%
    round())

@@

knitr::kable(temp_comp, digits = 1, format = 'markdown', caption = "Changes in deviance expained ($\\Delta$DE) and Akaike Information Criterion ($\\Delta$AIC) from including temperature in the models estimating probability of movement (P(M)) and speed when moving (Speed) as well as in species-level Hierarchical Resource Selection Functions (HRSF). Unlike conventional marginal AIC [@akaike_new_1974], The AIC values provided by the `mgcv` package for `R` are calculated using the models' estimated effective degrees of freedom and the Bayesian variance-covariance matrices, which avoids issues related to random effects or Restricted Maximum Likelihood (REML).")
```

@@

```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the estimated effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{\\text{st}}$ (\\textbf{A}) and day of year at 12:00 PDT (\\textbf{B}). Predictions in the surface plots extend to 10% of the range away from each datum. The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to $2^{\\pm 2}$ for ease of readability."}
# not specifying dt, or doy for A and time for B because change is relative
knitr::include_graphics('../figures/distance.png')
```

@@

```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each location's average distance in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.72 to 1.29). The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest cover."}
knitr::include_graphics('../figures/bc-displ-2100.png')
```

@@

```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in relative selection strength (RSS) for their current range, irrespective of RSS quantile or climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in RSS for the bottom 5\\%, median, and top 5\\% RSS values within the species' current extent. Changes are relative to the mean RSS for each quantile in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
knitr::include_graphics('../figures/rss-local-cc-predictions-quantiles.png')
```

```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.26 to 117). The increase in boreal caribou's RSS for high elevations at warmer temperatures was highly uncertain and often not credible at the 95\\% level."}
knitr::include_graphics('../figures/bc-rss-2100.png')
```

@@

\noindent We have demonstrated that temperature is an important determinant of whether, how, and where mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources). Disentangling temperature from other drivers is important for predicting how changes in climate will affect mammalian movement when other drivers remain approximately constant. However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are exposed to increasingly frequent conditions that are both extreme and novel. Predicting mammals' responses at the data-scarce fringes of the surface plots in Figs. \@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear. At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures. As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023]. For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [@mortlock_sleep_2024] and likelihood to enter torpor [@fjelldal_nightly_2021]. Such changes may alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024]. At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023], increase their chances of finding food or being predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021]. These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011].
